---
title: 'Stage 2: Exploratory Data Analysis'
author: "Isaac Slagel and Jack Welsh"
date: "4/18/2019"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 4.5
  html_document:
    df_print: paged
header-includes: \usepackage{float}
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(kableExtra)
library(knitr)
adoptions <- read_csv("adoptions.csv")
```

# EDA Main Report

## Introduction

  Our project plans to look into trend in animal treatment in Dallas animal shelters. Using a dataset containing information on all recent animals brought into the shelter, we are going to try to answer some of the following questions. What traits affect a dog's probability of adoption? Are there seasonal trends in the adoption of different animals? Are animals who enter the animal shelter with chips more likely to be returned to their owner? On the table below you can see some of the variables we are working with in the dataset.



## Dogs

  Dogs are the most common animal winding up in the animal shelters which form our dataset. From our total of 44,194 dogs, 35% were adopted, 30% were returned to owners, 16% were trasfered, and 15% were euthanized. The most common breed of dog in our dataset is the pitbull, with 10,033 being admited to animal shelters in between 10/01/2017 and 04/03/2019. We are especially interested in how pitbulls are treated within our datasets as these dogs have an infamous reputation of being overly agressive. Is it possible that this reputation results in a lower adoption rate for pitbulls that other breeds?

#### Pitbulls


  To look into how pitbulls are handled within animal shelters, we decided to explore how the outcomes of pitbulls may differ from non-pitbulls. Table 1 describes the differences in outcome rates for non-pitbulls compared to pitbulls.



```{r pitbull table, echo = FALSE}
proper=function(x) paste0(toupper(substr(x, 1, 1)), tolower(substring(x, 2)))

adoptions %>%
  filter(dog == 1) %>%
  mutate(outcome = as.factor(outcome_type),
         pitbull = as.factor(pitbull)) %>%
  group_by(pitbull, outcome_type) %>%
  dplyr::count()%>%
  rename("freq"=n)%>%
  group_by(pitbull)%>%
  mutate(freq = freq/ sum(freq))%>%
  spread(outcome_type,freq) %>%
  gather(key = Outcome, value = value, -pitbull) %>%
  spread(key = pitbull, value = value) %>%
  select(1,3,2) %>%
  rename( "Pitbull (%)" = `1`, "Non-Pitbull (%)" = `0`) %>%
  mutate(Outcome = proper(Outcome), `Pitbull (%)` =100*`Pitbull (%)` , `Non-Pitbull (%)`=`Non-Pitbull (%)`*100 )%>%
  arrange(desc(`Pitbull (%)`))%>%
  kable(digits = 2, format = "latex", booktabs = T, caption = "Pitbull Outcomes")%>%
  kable_styling(latex_options =  c("striped", "hold_position"))

```


  We see that pitbulls, while only adopted at a slightly lower rate (~5%), are euthanized at well over double the rate of other dogs. Further, we see that other dogs have a much higher chance of being trasfered to another facility or returned to their owner. It would be valuble to get a better understanding of this relationship, but we likely need to control for things like chip status and intake condition. Moving foward we would like to look into some binomial regression to look into the outcomes of pitbulls. To do this we plan to make a more general outcome variable of Dead or Alive status (to reduce the dimension of the current outcome_type variable). In fitting this model we plan to control for intake status (with which we still need to do some work on string analysis), chip_status, and month.

## Dogs vs Cats

  In Minnesota and Iowa, our home states, animal shelters often report having more difficulty dealing with stray cats. There are lots of problems that feral cats cause including environmental damage and spread of disease. We were interested if we can see these kinds of differences in our data. Do cats spend longer in animal shelters? Are they euthanized at higher rates? To describe some of these differences we looked we created a summary graph of different characteristics of dogs and cats. Since our dataset contains observations of 44194 dogs and only 12891 cats, we created this chart in terms of proportion of each animal matching a certain critera.


```{r DogCat Figure, echo = FALSE, fig.align='center', out.width = "60%", fig.pos='H', fig.cap="Cat vs Dog Characteristics"}
adoptions %>%
  filter(dog == 1 | cat == 1) %>%
  mutate(Species = ifelse(dog == 1, "Dog", "Cat")) %>%
  mutate(returned = ifelse(outcome_type == "RETURNED TO OWNER", 1, 0)) %>%
  group_by(Species) %>%
  summarise(count = n(),
            meanDays = mean(shelter_days, na.rm = TRUE),
            `Prop. Euthanized` = sum(euthanized)/count,
            `Prop. Adopted` = sum(adopted)/count,
            `Prop. Stray` = sum(stray)/count,
           `Prop. Returned` = sum(returned)/count,
           `Prop. Chipped` = sum(chip_status == "SCAN CHIP", na.rm = TRUE)/count)%>%
  gather(key = key, value = proportion, 4:8) %>%
  ggplot(aes(x=Species, y= proportion)) +
    geom_bar(stat = "identity")+
    facet_grid(cols = vars(key))+
  scale_y_continuous(breaks = seq(0, 1, by = .1))
```


Interestingly enough, we see that the difference between proportions of cats and dogs adopted is not very pronounced. A slightly lower percentage of cats are adopted and a slightly higher percent of cats are euthanized compared to dogs. However, we do see that very few cats entering humane societys have scannable chips, and an even lower proportion are returned to their owner. Additionally it seems appears that fewer cats are brought into the shelter as strays. We may use this data to do another binomial analysis to piece apart differences between cat and dog adoptions. We may do a similar thing as the pitbull analysis with the creation of a dead or alive outcome variable. We would want to consider controlling for in this analysis would be breed, intake condition,and chip status.

## Time of Year

Another question that we had in our Stage 1 report was what seasonal trends exists in adoptions. One could imagine a situation where more puppies are adopted in December or more rabbits are adopted in early spring. To explore this we did some ploting to see how many animals were adopted in each months.

```{r Month Figure, echo = FALSE, fig.align='center', out.width = "60%", fig.pos='H', fig.cap="Animal Adoptions by Month"}
intake <- adoptions %>%
  mutate(intake_month = month(intake_date, label = TRUE)) %>%
  ggplot(aes(x=intake_month))+
  geom_bar()+ggtitle("Intake Month")
outcome <- adoptions %>%
  filter(!is.na(outcome_date)) %>%
  mutate(outcome_month = month(outcome_date, label = TRUE)) %>%
  ggplot(aes(x=outcome_month))+
  geom_bar()+ ggtitle("Outcome Month")
gridExtra::grid.arrange(intake, outcome)
```

These trends surprised us. We see a large drop in adoptions during the summer but a spike in the fall, winter, and early spring.  These trends were true for both Dog and Cat adoptions (see appendix). For rabbits, many were adopted in February and July. While we are unable to put together a rigourus defense for why these trends are occuring, we see that we may need to control for it in our binomial analysis.

\pagebreak

# Annotated Appendix and References

##Variables

```{r variable table, echo = FALSE}
variables <- c("Variable Name", "Variable Role", "Variable Type", "Range of Values", "Units")

breed <- c("animal_breed", "explanatory", "categorical", "296 unique breeds", "NA")
origin <- c("animal_origin", "explanatory", "categorical", "4 sources of shelter animals", "NA")
type <- c("animal_type", "explanatory", "categorical", "5 species of animal", "NA")
chip <- c("chip_status", "explanatory", "bianary", "(0,1)", "NA")

intake_type <- c("intake_type", "potential confounder", "categorial", "how animal came to be at the shelter", "NA")
outcome_type <- c("outcome_type", "reponse", "categorial", "how animals was removed from shelter", "NA")

intake_condition <- c("intake_condition", "potential confounder", "categorial", "keyword description of animal status at intake", "NA")
outcome_condition <- c("outcome_condition", "potential confounder", "categorial", "keyword description of animal status at outcome", "NA")

intake_date <- c("intake_date", "response", "date", "(2017-10-01, 2019-04-03)", "y-m-d")
outcome_date <- c("outcome_date", "response", "date", "(2017-10-01, 2019-04-03)", "y-m-d")

tab <- rbind(breed, origin, type, chip, intake_type, outcome_type, intake_condition, outcome_condition, intake_date, outcome_date)

colnames(tab) <- variables
rownames(tab) <- NULL

kable(tab, format = "latex", booktabs = T , caption = "Description of Variables") %>%
  column_spec( 4, width = "4cm") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"))

```


## Citations

1. [Lepper, M., Kass, P. H., & Hart, L. A. (2002). Prediction of adoption versus euthanasia among dogs and cats in a California animal shelter. Journal of Applied Animal Welfare Science, 5(1), 29-42.](https://www.tandfonline.com/doi/abs/10.1207/S15327604JAWS0501_3)

2. [Posage, J. M., Bartlett, P. C., & Thomas, D. K. (1998). Determining factors for successful adoption of dogs from an animal shelter. Journal of the American Veterinary Medical Association, 213(4), 478-482.](https://europepmc.org/abstract/med/9713528)

3. [Lampe, R., & Witte, T. H. (2015). Speed of dog adoption: Impact of online photo traits. Journal of applied animal welfare science, 18(4), 343-354.](https://www.tandfonline.com/doi/abs/10.1080/10888705.2014.982796)

## Exploring animal_type

```{r}
# animal type
adoptions %>%
  group_by(animal_type) %>%
  count(sort = TRUE)

adoptions %>%
  filter(!animal_type%in%c("D", "LIVESTOCK")) %>%
  ggplot(aes(x=animal_type))+
  geom_bar()
```

We see that most observations in our dataaset were of dogs, followed by cats. It seems there is 1 mislabeled dog which we will take care of in our data cleaning file.

## Exploring animal_breed

```{r}
## Which dog breeds are in our data?
adoptions %>%
  filter(animal_type == "DOG") %>%
  mutate(animal_breed = fct_lump(animal_breed, n=15)) %>%
  group_by(animal_breed) %>%
  count(sort = TRUE)

adoptions %>%
  filter(animal_type == "DOG") %>%
  mutate(animal_breed = fct_lump(animal_breed, n=15)) %>%
  group_by(animal_breed) %>%
  count(sort = TRUE)%>%
  ggplot(aes(x=fct_reorder(animal_breed,n), y=n))+
  geom_bar(stat="identity")+
  coord_flip()
```

We have a huge variety of dog breeds in our data. The pmost common breed is the pitbull, which we are planning to use for our analysis. There are many other dog breeds that are seen as agressive (rottweilers and boxers) that we may make into a new variable `aggressive`.

```{r}
## Which cat breeds are in our data?
adoptions %>%
  filter(animal_type == "CAT") %>%
  mutate(animal_breed = fct_lump(animal_breed, n=10)) %>%
  group_by(animal_breed) %>%
  count(sort = TRUE)

adoptions %>%
  filter(animal_type == "CAT") %>%
  mutate(animal_breed = fct_lump(animal_breed, n=15)) %>%
  group_by(animal_breed) %>%
  count(sort = TRUE)%>%
  ggplot(aes(x=fct_reorder(animal_breed,n), y=n))+
  geom_bar(stat="identity")+
  coord_flip()
```

Almost all of the cats in our data were domestic short hair cats.


```{r}
## Which types of wildlife are in our data?
adoptions %>%
  filter(animal_type == "WILDLIFE") %>%
  mutate(animal_breed = fct_lump(animal_breed, n=10)) %>%
  group_by(animal_breed) %>%
  count(sort = TRUE)
```

This variable is more for fun. Look at all these fun creatures they get to deal with at the animal shelter. We are kind of interested in rabbits so we will talk about that later.

## Exploring animal origin

```{r}
adoptions %>%
  filter(dog == 1 | cat == 1) %>%
  filter(!is.na(animal_origin)) %>%
  ggplot(aes(x=animal_origin))+
  geom_bar()+
  facet_grid(~animal_type)
```

Dogs and cats can come from a 3 possible origins: field, over the counter, and sweep. As of this moment, we are not sure what the difference between field and sweep are. However, it is interesting that only dogs are brought in from sweeps.  

## Exploring chip status

```{r}
## General distribution of variable
adoptions %>%
  filter(dog == 1 | cat == 1) %>%
  ggplot(aes(x=chip_status))+
  geom_bar()+
  facet_grid(~animal_type)

## Relationship with returned to owner status
adoptions %>%
  filter(dog == 1 | cat == 1) %>%
  filter(!is.na(chip_status)) %>%
  mutate(returned = as.factor(ifelse(outcome_type== "RETURNED TO OWNER", 1, 0))) %>%
  ggplot(aes(x=chip_status, fill = returned))+
  geom_bar(position = "fill")+
  facet_grid(~animal_type)
```

We see that most cats and dogs do not have chips. However, a larger proportion of dogs have chips than cats.

A higher proportion of dogs are returned to owners than cats in all three categories of chip status.

## Exploring intake_subtype

```{r}
## Which dog breeds are in our data?
adoptions %>%
  mutate(intake_subtype = fct_lump(intake_subtype, n=10)) %>%
  group_by(intake_subtype) %>%
  count(sort = TRUE)
```

This general intake listing does not give us a lot of infomration. However we do see that a few animals may have been returned (non-independent observation), and some animals are requested to be euthanized upon admission.

## Exploring intake_date and outcome_date

```{r}
intake <- adoptions %>%
  mutate(intake_month = month(intake_date, label = TRUE)) %>%
  ggplot(aes(x=intake_month))+
  geom_bar()+
  ggtitle("Intake Month")

outcome <- adoptions %>%
  filter(!is.na(outcome_date)) %>%
  mutate(outcome_month = month(outcome_date, label = TRUE)) %>%
  ggplot(aes(x=outcome_month))+
  geom_bar()+ ggtitle("Outcome Month")
gridExtra::grid.arrange(intake, outcome)
```

```{r}
# Ammount of time spent at the humane society looks to be about the same from month to month.

adoptions %>%
  mutate(diff_time=difftime(outcome_date, intake_date))%>%
  select(diff_time, month)%>%
  ggplot(aes(x=diff_time))+
  geom_density()+
  facet_wrap(~as.factor(month))


```

### Differences in income day

```{r}
adoptions%>%
  mutate(out_day=day(outcome_date))%>%
  ggplot(aes(x=out_day))+
  geom_bar()+ ggtitle("Outcome day")


adoptions%>%
  mutate(in_day=day(intake_date))%>%
  ggplot(aes(x=in_day))+
  geom_bar()+ ggtitle("Outcome day")

```




## Pitbulls and Adoptions
```{r}
adoptions %>%
  filter(dog == 1) %>%
  mutate(adopted = as.factor(adopted),
         pitbull = as.factor(pitbull)) %>%
  ggplot(aes(x = pitbull, fill = adopted))+
  geom_bar(position = "fill")
```

There does not seem to be a large difference between the adoption rates of pitbull and non-pitbulls. We should try and filter out some of the non-adopted dogs though, not all dogs come to the shelter to be adopted.

```{r}
adoptions %>%
  filter(dog == 1) %>%
  mutate(euthanized = as.factor(euthanized),
         pitbull = as.factor(pitbull)) %>%
  ggplot(aes(x = pitbull, fill = euthanized))+
  geom_bar(position = "fill")

```

While pitbulls are not adopted at a higher rate, the are euthanized at a higher rate than other dogs. Lets look at all possible outcomes to see how this is possible.

```{r}
adoptions %>%
  filter(dog == 1) %>%
  mutate(outcome = as.factor(outcome_type),
         pitbull = as.factor(pitbull)) %>%
  ggplot(aes(x = pitbull, fill = outcome))+
  geom_bar(position = "fill")


adoptions %>%
  filter(dog == 1) %>%
  mutate(outcome = as.factor(outcome_type),
         pitbull = as.factor(pitbull)) %>%
  group_by(pitbull, outcome_type) %>%
  count() %>%
  rename("freq"=n)%>%
  group_by(pitbull)%>%
  mutate(freq = freq/ sum(freq))%>%
  spread(outcome_type,freq)
```

A far higher percentage of pitbulls are euthanized than other breeds. Other dogs are adopted, returned to owners, and trasfered at a higher rate.

## Differences between dogs and cats

```{r}
adoptions %>%
  filter(dog == 1 | cat == 1) %>%
  mutate(Species = ifelse(dog == 1, "Dog", "Cat")) %>%
  mutate(returned = ifelse(outcome_type == "RETURNED TO OWNER", 1, 0)) %>%
  group_by(Species) %>%
  summarise(count = n(),
            meanDays = mean(shelter_days, na.rm = TRUE),
            `Proportion Euthanized` = sum(euthanized)/count,
            `Proportion Adopted` = sum(adopted)/count,
            `Proportion Stray` = sum(stray)/count,
           `Proportion Returned` = sum(returned)/count,
           `Proportion Chipped` = sum(chip_status == "SCAN CHIP", na.rm = TRUE)/count)%>%
  gather(key = key, value = proportion, 4:8) %>%
  ggplot(aes(x=Species, y= proportion)) +
    geom_bar(stat = "identity")+
    facet_grid(cols = vars(key))

```

Cats are returned at a much lower rate than dogs. Cats have a much lower proportion with scannable chips than dogs.

```{r}
adoptions %>%
  filter(dog == 1 | cat == 1) %>%
  filter(shelter_days < 25) %>%
  mutate(DogCat = ifelse(dog == 1, "Dog", "Cat")) %>%
  ggplot(aes(x=shelter_days, y=..density..)) +
  geom_histogram()+
  facet_grid(~DogCat)
```

Dogs and cats show similar trends in the number of days until adoptions.


## What about rabbits?

```{r}
adoptions  %>%
  filter(animal_breed== "RABBIT SH") %>%
  filter(adopted == 1) %>%
  mutate(month = lubridate::month(outcome_date, label = TRUE)) %>%
  ggplot(aes(x=month)) +
  geom_bar()
```

Alright, so maybe it is not the case that more rabits are adopted over easter. Are there trends in dog adoptions?

```{r}
adoptions  %>%
  filter(dog == 1 | cat == 1) %>%
  mutate(DogCat = ifelse(dog == 1, "Dog", "Cat")) %>%
  filter(adopted == 1, !is.na(month)) %>%
  mutate(month = lubridate::month(outcome_date, label = TRUE)) %>%
  ggplot(aes(x=month, y = ..prop.., group = 1)) +
  geom_bar()+
  facet_grid(rows = vars(DogCat), switch = "both")
```

These both seem to follow the trend of fewer adoptions in the summer.

## Intake Condition

We are doing a preliminar investigation into this. There is no domunetation on what exactly all the different levels mean. For example we do not know what the difference between a manageable animal and a treatable animal are or what the differnce between a normal and a healthy animal are. 

```{r}
adoptions_expanded_intake=adoptions%>%
  mutate(healthy_intake=ifelse(grepl("^HEALTHY.*",intake_condition),1,0),
         contagious_intake=ifelse(grepl(".*[^NON-]CONTAGIOUS", intake_condition),1,0),
         untreatable_intake=ifelse(grepl(".*(UNTREATABLE).*", intake_condition),1,0),
         treatable_intake=ifelse(grepl("^TREATABLE.*", intake_condition),1,0),
         manageable_intake=ifelse(grepl(".*MANAGEABLE.*", intake_condition),1,0),
         rehabitable_intake=ifelse(grepl(".*REHABILITABLE.*", intake_condition),1,0),
         normal_intake=ifelse(grepl(".*NORMAL.*", intake_condition),1,0))
```

## Healthy

It looks like a slightly higher proportion of cats are healthy (0.0314) than dogs (0.0131), however both of there are suprisingly low. 

```{r}
adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(Healthy=mean(healthy_intake))

adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(Healthy=mean(healthy_intake))%>%
  ggplot(aes(x=animal_type, y=Healthy))+
  geom_bar(stat="identity")
```

## Contagious

Even though a larger proportion of cats are found or given up healthy, a larger proportion of cats are contagious on intage (cats 0.0565 and dogs 0.0159).

```{r}
adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(contagious=mean(contagious_intake))%>%
  ggplot(aes(x=animal_type, y=contagious))+
  geom_bar(stat="identity")

adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(contagious=mean(contagious_intake))

```

##Treatable

It looks like there is little difference in the proportion of cats and dogs that on intake are treatable. 

```{r}
adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(treatable=mean(treatable_intake))%>%
  ggplot(aes(x=animal_type, y=treatable))+
  geom_bar(stat="identity")

adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(treatable=mean(treatable_intake))

```


## Manageable

It appears that these are about the same for both cats and dogs 

```{r}
adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(manageable=mean(manageable_intake))%>%
  ggplot(aes(x=animal_type, y=manageable))+
  geom_bar(stat="identity")

adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(manageable=mean(manageable_intake))

```



## Rehabilatable

It appears that these are about the same for both cats and dogs.

```{r}
adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(rehab=mean(rehabitable_intake))%>%
  ggplot(aes(x=animal_type, y=rehab))+
  geom_bar(stat="identity")

adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(rehabitable=mean(rehabitable_intake))
```

# Normal

It looks like this was only used a few times and most likely is not important. 

```{r}
adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(normal=mean(normal_intake))%>%
  ggplot(aes(x=animal_type, y=normal))+
  geom_bar(stat="identity")

adoptions_expanded_intake%>%
  filter(cat==1|dog==1)%>%
  group_by(animal_type)%>%
  summarise(normal=mean(normal_intake))
```


